\section{Model and Assumptions}\label{sec:model}
% \newcommand{\Expect}{E}
\newcommand{\Prob}{P}
\newcommand{\F}{F}
\newcommand{\Fhat}{{\hat{\F}}}
\newcommand{\qhat}{{\hat{q}}}
\newcommand{\Sx}{{\mathcal{S}_X}}
\newcommand{\Sr}{{\mathcal{S}_R}}
\newcommand{\Sn}{s_n}
\newcommand{\urange}{u_{\mbox{\footnotesize range}}}
%\newcommand{\umin}{u_{\mbox{\footnotesize min}}}
\newcommand{\qhatOp}{{\bold{\qhat}}}


%\newcommand{\EUFq}{\mbox{EU($q;\F$)}}
%\newcommand{\EUFhqh}{\mbox{EU($\qhat;\Fhat$)}}
%\newcommand{\EUFqh}{\mbox{EU($\qhat;\F$)}}
%\newcommand{\CEFq}{\mbox{CE($q;\F$)}}
%\newcommand{\CEFhqh}{\mbox{CE($\qhat;\Fhat$)}}
%\newcommand{\CEFqh}{\mbox{CE($\qhat;\F$)}}





We consider a classical financial portfolio selection problem involving a risky asset with
random return rate $R$ and a risk-free asset with return rate of $0\%$ for simplicity. We
also suppose that the investor's risk aversion can be characterized using expected utility
theory using a strictly increasing concave utility function $u$, and that the investor has
access to side information regarding the returns. This information might be the result of
processing the most recent financial or economic news, etc. We let this information be
described as a vector of $p$ normalized random features $[X_1,X_2,\dots,X_p]$. In this
context, if the the distribution $\F$ of the pair $(X,R)$ of side information and return
is known, a linear investment policy that exploits the side information optimally for this
investor can be obtained by solving the following optimization problem:
\begin{equation}
\maximize_{q\in\Re^p}\;\;\;\E_\F[u(R\cdot q^TX)]\;,\label{EUF}
\end{equation}
where it is assumed that short-selling is permitted. 

In practice however, the exact distribution describing the relation between $X$ and $R$ is
not available at the time of designing the investment policy and one might instead need to
exploit a sample set $\Sn:=\{(x_i,r_i)\}_{i=1}^n$ drawn independently and identically from
$\F$. Unfortunately, when the sample size $n$ is relatively small compared to $p$, it is
well known that the problem \eqref{EUF} using the empirical distribution $\Fhat$ obtained
from sample $s_n$ can suffer from severe overfitting and produce investment policies that
perform badly out of sample. This is for instance illustrated in the following example.

\begin{ex}
  Consider for instance a case where $n=p$ and each feature $X_i$ is independantly and
  identically drawn from a Gaussian distribution. Given that it is well known that the
  probability that the random matrix $\Xi := [X_1\;X_2\;\dots\;X_n]$ be singular is null,
  then one can easily establish that problem \eqref{EUF} with $\Fhat$ is
  unbounded. Indeed, one can verify that $r_i \bar{q}^T x_i = 1$ for all $i=1,\dots,n$
  when $\bar{q}$ is set to ${\Xi^{-1}}^T [1/r_1\;1/r_2\;\dots\;1/r_n]^T$. Hence, one can
  achieve an arbitrarily large empirical expected utility by investing according to
  $\alpha\bar{q}$ for $\alpha>0$.
\end{ex}

To prevent issues associated to overfitting, one might instead seek the optimal solution
of the following regularized empirical expected utility maximization problem:
\begin{equation}
\maximize_{q\in\Re^p}\;\;\;\E_\Fhat[u(R\cdot q^TX)]+\lambda\|q\|^2\;.\label{EUFhatReg}
\end{equation}
Note that when it exists, we will refer to the optimal solution of this problem as $\qhat$.



\input{oos_performance}
\input{true_optimal}
\input{big_data}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "big_data_portfolio_optimization"
%%% End:
